---
title: "Advanced topics: morphisms and complex models"
author: "Simon Frost"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Advanced topics: morphisms and complex models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

This vignette covers advanced features of algebraicodin:

- **Morphisms** between Petri nets (typed nets as structure-preserving maps)
- **Schema migration** for transforming models
- **Complex compositions** combining composition and stratification
- **Staged infections** (linear chain trick)
- **Multi-component models** (SIRH with hospitalization and death)

These features demonstrate how category theory provides a principled
framework for building and transforming epidemiological models.

```{r setup}
library(algebraicodin)
```

## Morphisms of Petri nets

A **morphism** $\phi: P_1 \to P_2$ between Petri nets maps species to
species and transitions to transitions in a structure-preserving way. In
algebraicodin, this is exactly what `typed_petri()` creates—a morphism
from a concrete model to a type system.

The morphism must preserve the incidence structure: if species $s$ is an
input to transition $t$ in $P_1$, then $\phi(s)$ must be an input to
$\phi(t)$ in $P_2$.

```{r morphism-example}
ont <- infectious_ontology()

sir <- labelled_petri_net(
  c("S", "I", "R"),
  inf = c("S", "I") %=>% c("I", "I"),
  rec = "I" %=>% "R"
)
phi <- typed_petri(sir, ont,
  species_types    = c(S = "Pop", I = "Pop", R = "Pop"),
  transition_types = c(inf = "infect", rec = "disease")
)
```

The morphism $\phi$ encodes:

| Component | SIR (domain) | Ontology (codomain) |
|-----------|-------------|---------------------|
| Species | S, I, R | Pop, Pop, Pop |
| Transitions | inf, rec | infect, disease |
| Input arcs | S→inf, I→inf, I→rec | Pop→infect, Pop→infect, Pop→disease |
| Output arcs | I→inf, I→inf, R→rec | Pop→infect, Pop→infect, Pop→disease |

```{r morphism-maps}
cat("Species map:", phi@species_map, "\n")
cat("Transition map:", phi@transition_map, "\n")
cat("Input arc map:", phi@input_map, "\n")
cat("Output arc map:", phi@output_map, "\n")
```

```{r morphism-diagram}
plot_petri(phi)
```

## Schema migration via catlab

The catlab package provides `delta_migrate()` for pullback migration
along schema functors. This can transform data between different
representations of the same mathematical structure.

```{r migration}
# Two schemas for the same data, related by a functor
SchA <- acsets::BasicSchema(
  obs = c("Person", "City"),
  homs = list(acsets::hom("lives_in", "Person", "City"))
)
SchB <- acsets::BasicSchema(
  obs = c("Individual", "Location"),
  homs = list(acsets::hom("residence", "Individual", "Location"))
)

# Functor F: SchA → SchB
F <- catlab::FinFunctor(
  ob_map  = list(Person = "Individual", City = "Location"),
  hom_map = list(lives_in = "residence"),
  dom = catlab::FinCat(schema = SchA),
  codom = catlab::FinCat(schema = SchB)
)

# Data on SchB
DataB <- acsets::acset_type(SchB)
db <- DataB()
acsets::add_part(db, "Individual")
acsets::add_part(db, "Individual")
acsets::add_part(db, "Location")
acsets::set_subpart(db, 1L, "residence", 1L)
acsets::set_subpart(db, 2L, "residence", 1L)

# Migrate to SchA
DataA <- acsets::acset_type(SchA)
da <- catlab::delta_migrate(F, db, DataA)
cat("Persons:", acsets::nparts(da, "Person"), "\n")
cat("Cities:", acsets::nparts(da, "City"), "\n")
cat("Person 1 lives_in:", acsets::subpart(da, 1L, "lives_in"), "\n")
```

This migration preserves the relational structure while renaming the
components—analogous to renaming compartments in an epidemiological
model.

## Staged infection (linear chain trick)

The linear chain trick models non-exponential residence times by
splitting a compartment into multiple stages. An infectious period with
Erlang-distributed duration uses $k$ sequential I-stages.

```{r stages}
# SIR with 3 infectious stages
sir_stages <- labelled_petri_net(
  c("S", "I1", "I2", "I3", "R"),
  inf1 = c("S", "I1") %=>% c("I1", "I1"),
  inf2 = c("S", "I2") %=>% c("I2", "I2"),
  inf3 = c("S", "I3") %=>% c("I3", "I3"),
  prog12 = "I1" %=>% "I2",
  prog23 = "I2" %=>% "I3",
  rec    = "I3" %=>% "R"
)
species_names(sir_stages)
transition_names(sir_stages)
```

```{r stages-diagram}
plot_petri(sir_stages)
```

```{r stages-odin}
cat(pn_to_odin(sir_stages, "ode"))
```

### Programmatic stage generation

For arbitrary numbers of stages, build the Petri net programmatically:

```{r stages-programmatic}
make_staged_sir <- function(n_stages) {
  snames <- c("S", paste0("I", seq_len(n_stages)), "R")
  pn <- LabelledPetriNet()
  sids <- setNames(
    vapply(snames, function(s) acsets::add_part(pn, "S", sname = s), integer(1)),
    snames
  )

  # Infection from each stage

  for (k in seq_len(n_stages)) {
    ik <- paste0("I", k)
    tname <- paste0("inf", k)
    tid <- acsets::add_part(pn, "T", tname = tname)
    acsets::add_part(pn, "I", is = sids[["S"]], it = tid)
    acsets::add_part(pn, "I", is = sids[[ik]], it = tid)
    acsets::add_part(pn, "O", os = sids[[ik]], ot = tid)
    acsets::add_part(pn, "O", os = sids[[ik]], ot = tid)
  }

  # Progression between stages
  for (k in seq_len(n_stages - 1L)) {
    from <- paste0("I", k)
    to <- paste0("I", k + 1L)
    tname <- paste0("prog", k, k + 1)
    tid <- acsets::add_part(pn, "T", tname = tname)
    acsets::add_part(pn, "I", is = sids[[from]], it = tid)
    acsets::add_part(pn, "O", os = sids[[to]], ot = tid)
  }

  # Recovery from last stage
  tid <- acsets::add_part(pn, "T", tname = "rec")
  acsets::add_part(pn, "I", is = sids[[paste0("I", n_stages)]], it = tid)
  acsets::add_part(pn, "O", os = sids[["R"]], ot = tid)

  pn
}

sir5 <- make_staged_sir(5)
cat("5-stage SIR: ", acsets::nparts(sir5, "S"), "species,",
    acsets::nparts(sir5, "T"), "transitions\n")
```

```{r stages-run, fig.width = 6, fig.height = 4}
code5 <- pn_to_odin(sir5, "ode")
gen5 <- odin2::odin(code5)
pars5 <- c(
  setNames(as.list(rep(0.0005, 5)), paste0("inf", 1:5)),
  setNames(as.list(rep(1.0, 4)), paste0("prog", 1:4, 2:5)),
  list(rec = 1.0, S0 = 990, I10 = 10),
  setNames(as.list(rep(0, 4)), paste0("I", 2:5, "0")),
  list(R0 = 0)
)
sys5 <- dust2::dust_system_create(gen5, pars5, n_particles = 1)
dust2::dust_system_set_state_initial(sys5)
t <- seq(0, 40, by = 0.1)
y5 <- dust2::dust_system_simulate(sys5, t)

# Compare 1-stage vs 5-stage
gen1 <- odin2::odin(pn_to_odin(sir, "ode"))
sys1 <- dust2::dust_system_create(gen1,
  list(inf = 0.0005, rec = 0.25, S0 = 990, I0 = 10, R0 = 0),
  n_particles = 1)
dust2::dust_system_set_state_initial(sys1)
y1 <- dust2::dust_system_simulate(sys1, t)

library(ggplot2)
df <- data.frame(
  time = rep(t, 2),
  I = c(y1[2, ], rowSums(t(y5[2:6, , drop = FALSE]))),
  model = rep(c("Exponential (1 stage)", "Erlang (5 stages)"), each = length(t))
)
ggplot(df, aes(time, I, colour = model)) +
  geom_line(linewidth = 0.8) +
  labs(title = "Effect of staged infection on epidemic dynamics",
       x = "Time", y = "Total infectious") +
  theme_minimal()
```

The 5-stage model produces a sharper, later epidemic peak because the
Erlang distribution has less variance than the exponential.

## SIRH: hospitalization and death

Complex models combine multiple transition types. Here we build SIRH
(with hospitalization and death) compositionally:

```{r sirh}
infection   <- exposure_petri("S", "I", "I", "inf")
recovery    <- spontaneous_petri("I", "R", "rec")
hosp        <- spontaneous_petri("I", "H", "hosp")
hosp_rec    <- spontaneous_petri("H", "R", "hosp_rec")
death_inf   <- spontaneous_petri("I", "D", "death_I")
death_hosp  <- spontaneous_petri("H", "D", "death_H")

w <- catlab::uwd(
  outer      = c("S", "I", "R", "H", "D"),
  infection  = c("S", "I"),
  recovery   = c("I", "R"),
  hosp       = c("I", "H"),
  hosp_rec   = c("H", "R"),
  death_inf  = c("I", "D"),
  death_hosp = c("H", "D")
)
sirh <- oapply(w, list(infection, recovery, hosp, hosp_rec, death_inf, death_hosp))
cat(pn_to_odin(apex(sirh), "ode"))
```

```{r sirh-diagram, fig.show = "hold"}
plot_uwd(w)
plot_petri(sirh)
```

```{r sirh-run, fig.width = 6, fig.height = 4}
gen <- odin2::odin(pn_to_odin(apex(sirh), "ode"))
pars <- list(
  inf = 0.0005, rec = 0.2, hosp = 0.05, hosp_rec = 0.1,
  death_I = 0.01, death_H = 0.02,
  S0 = 990, I0 = 10, R0 = 0, H0 = 0, D0 = 0
)
sys <- dust2::dust_system_create(gen, pars, n_particles = 1)
dust2::dust_system_set_state_initial(sys)
y <- dust2::dust_system_simulate(sys, t)

snames <- species_names(apex(sirh))
df <- data.frame(
  time = rep(t, length(snames)),
  value = as.vector(t(y)),
  compartment = factor(rep(snames, each = length(t)), levels = snames)
)
ggplot(df, aes(time, value, colour = compartment)) +
  geom_line(linewidth = 0.7) +
  labs(title = "SIRH model with death",
       x = "Time", y = "Population") +
  theme_minimal()
```

## Combining composition and stratification

The real power emerges when we combine composition and stratification.
Here we compose an SEIR model, then stratify it by age:

```{r seir-age}
# 1. Build SEIR compositionally
exposure    <- exposure_petri("S", "I", "E", "exp")
progression <- spontaneous_petri("E", "I", "prog")
recovery    <- spontaneous_petri("I", "R", "rec")

w_seir <- catlab::uwd(
  outer       = c("S", "E", "I", "R"),
  exposure    = c("S", "I", "E"),
  progression = c("E", "I"),
  recovery    = c("I", "R")
)
seir <- oapply(w_seir, list(exposure, progression, recovery))
seir_pn <- apex(seir)

# 2. Type the SEIR model
ont <- infectious_ontology()
seir_typed <- typed_petri(seir_pn, ont,
  species_types    = c(S = "Pop", E = "Pop", I = "Pop", R = "Pop"),
  transition_types = c(exp = "infect", prog = "disease", rec = "disease")
)
seir_aug <- add_reflexives(seir_typed, list(
  S = "strata", E = "strata", I = "strata", R = "strata"
))

# 3. Define 3 age groups
age3 <- labelled_petri_net(
  c("Ch", "Ad", "El"),
  aging_CA = "Ch" %=>% "Ad",
  aging_AE = "Ad" %=>% "El"
)
age3_typed <- typed_petri(age3, ont,
  species_types    = c(Ch = "Pop", Ad = "Pop", El = "Pop"),
  transition_types = c(aging_CA = "strata", aging_AE = "strata")
)
age3_aug <- add_reflexives(age3_typed, list(
  Ch = c("infect", "disease"),
  Ad = c("infect", "disease"),
  El = c("infect", "disease")
))

# 4. Stratify
seir_age <- seir_aug %x% age3_aug
cat("Species:", length(species_names(seir_age)), "\n")
cat(paste(species_names(seir_age), collapse = ", "), "\n\n")
cat("Transitions:", length(transition_names(seir_age)), "\n")
cat(paste(transition_names(seir_age), collapse = ", "), "\n")
```

```{r seir-age-diagram}
plot_petri(seir_age)
```

This produces a 12-species, 17-transition model from a 3-line
composition and a 3-line stratification specification.

```{r seir-age-code}
cat(pn_to_odin(seir_age, "ode"))
```

## Performance note

The algebraic overhead is purely at model construction time (R-level
manipulation of ACSet data structures). Once `pn_to_odin()` generates the
code and odin2 compiles it to C++, simulation performance is identical to
hand-written models.

For a 12-species model, construction takes < 1 ms:

```{r benchmark}
system.time({
  for (i in 1:100) {
    typed_product(seir_aug, age3_aug)
  }
})
```

## Summary

| Technique | Use case | Functions |
|-----------|----------|-----------|
| **Morphisms** | Type systems, ontologies | `typed_petri()`, `flatten()` |
| **Migration** | Schema transformation | `catlab::delta_migrate()` |
| **Staged infection** | Non-exponential distributions | Programmatic PN construction |
| **Multi-component** | Complex disease models | `oapply()`, `compose()` |
| **Composition + stratification** | Modular, scaled models | `%o%` + `%x%` |
