---
title: "Getting started with algebraicodin"
author: "Simon Frost"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Getting started with algebraicodin}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

**algebraicodin** uses applied category theory to build epidemiological
models compositionally, then generates [odin2](https://mrc-ide.github.io/odin2/)
code for high-performance simulation via
[dust2](https://mrc-ide.github.io/dust2/). Instead of writing ODE
equations by hand, you define a *Petri net*—a bipartite graph of species
(compartments) and transitions (processes)—and the package derives the
equations automatically.

This vignette walks through a basic SIR (Susceptible–Infectious–Recovered)
model, comparing the algebraic approach with hand-written odin2 code.

```{r setup}
library(algebraicodin)
```

## Defining a Petri net

A **labelled Petri net** describes a reaction network with named species
and named transitions. Each transition has input arcs (consumed species)
and output arcs (produced species).

For an SIR model, we have:

- **Species**: S (susceptible), I (infectious), R (recovered)
- **Transitions**:
  - `inf`: S + I → I + I (infection: one S and one I produce two I)
  - `rec`: I → R (recovery)

```{r sir-petri}
sir <- labelled_petri_net(
  c("S", "I", "R"),
  inf = c("S", "I") %=>% c("I", "I"),
  rec = "I" %=>% "R"
)
```

We can inspect the structure:

```{r sir-inspect}
species_names(sir)
transition_names(sir)
transition_matrices(sir)
```

The **stoichiometry matrix** shows the net change in each species for each
transition. For infection, S decreases by 1 and I increases by 1; for
recovery, I decreases by 1 and R increases by 1.

## Visualising the Petri net

The `plot_petri()` function renders the Petri net as a bipartite graph
via [DiagrammeR](https://rich-iannone.github.io/DiagrammeR/). Species
appear as blue circles and transitions as orange boxes:

```{r sir-diagram}
plot_petri(sir)
```

## Generating odin2 code

The `pn_to_odin()` function translates the Petri net into odin2 DSL code.
The default type is `"ode"` (continuous deterministic):
 
```{r sir-odin}
sir_code <- pn_to_odin(sir, type = "ode")
cat(sir_code)
```

This generates mass-action kinetics where each transition rate is the
product of the rate constant and the concentrations of input species.

## Comparison with hand-written odin2

Here is the equivalent model written directly in the odin2 DSL:

```{r sir-handwritten}
sir_handwritten <- "
## Parameters
inf <- parameter()
rec <- parameter()

## Initial conditions
S0 <- parameter()
I0 <- parameter()
R0 <- parameter()
initial(S) <- S0
initial(I) <- I0
initial(R) <- R0

## Mass-action rates
rate_inf <- inf * S * I
rate_rec <- rec * I

## Derivatives
deriv(S) <- -rate_inf
deriv(I) <- rate_inf - rate_rec
deriv(R) <- rate_rec
"
```

The generated code is structurally identical to the hand-written version.
The algebraic approach guarantees that the stoichiometry is internally
consistent—every molecule consumed by a transition is accounted for in the
outputs.

## Running the model

We can compile and run the generated odin2 code using dust2:

```{r sir-run}
gen <- odin2::odin(sir_code)
pars <- list(inf = 0.0005, rec = 0.25, S0 = 990, I0 = 10, R0 = 0)
sys <- dust2::dust_system_create(gen, pars, n_particles = 1)
dust2::dust_system_set_state_initial(sys)
t <- seq(0, 40, by = 0.1)
y <- dust2::dust_system_simulate(sys, t)
```

```{r sir-plot, fig.width = 6, fig.height = 4}
df <- data.frame(
  time = rep(t, 3),
  value = c(y[1, ], y[2, ], y[3, ]),
  compartment = rep(c("S", "I", "R"), each = length(t))
)
library(ggplot2)
ggplot(df, aes(time, value, colour = compartment)) +
  geom_line(linewidth = 0.8) +
  labs(title = "SIR model (algebraicodin → odin2)",
       x = "Time", y = "Population", colour = "Compartment") +
  theme_minimal()
```

## Using deSolve directly

For quick exploration without C++ compilation, the `vectorfield()`
function generates a pure-R ODE function compatible with
[deSolve](https://CRAN.R-project.org/package=deSolve):

```{r sir-desolve}
library(deSolve)
vf <- vectorfield(sir)
state <- c(S = 990, I = 10, R = 0)
parms <- c(inf = 0.0005, rec = 0.25)
out <- ode(state, t, vf, parms)
head(out)
```

## Summary

| Approach | Pros | Cons |
|----------|------|------|
| **algebraicodin** | Compositional; automatic stoichiometry; multi-target | Abstraction overhead for simple models |
| **Hand-written odin2** | Direct control; familiar syntax | Manual bookkeeping; error-prone for large models |
| **deSolve** | No compilation needed; immediate | Slower for large simulations; pure R |

For small models, the approaches are essentially equivalent. The power of
algebraicodin becomes clear when composing models from reusable parts
(see `vignette("composition")`) or stratifying by age or risk groups
(see `vignette("stratification")`).
