---
title: "Array-based code generation"
author: "Simon Frost"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Array-based code generation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6
)
set.seed(42)
```

## Introduction

When stratifying epidemiological models (by age group, location, strain, etc.),
the number of state variables and equations grows multiplicatively with the number
of strata. For a basic SIR model with 3 compartments, stratifying by 10 age
groups produces 30 state variables and dozens of flow equations in the scalar
(expanded) representation.

The **array-based code generator** `sf_to_odin_array()` addresses this by
producing compact odin2 code that uses array indexing. Instead of generating one
equation per compartment per stratum, it produces a single array-indexed equation
per base compartment, with odin2 expanding the loops internally. This makes the
generated code:

- **Compact** — code size grows O(1) rather than O(n) with strata count
- **Readable** — structure mirrors the un-stratified model
- **Scalable** — 50+ strata are practical without code bloat

```{r setup}
library(algebraicodin)
library(odin2)
library(dust2)
```

## Basic example: Age-stratified SIR

We start with a base SIR stock-flow model and stratify it by age group.

### Define the base model

```{r base-sir}
sir <- stock_and_flow(
  stocks = c("S", "I", "R"),
  flows = list(
    infection = flow(from = "S", to = "I", rate = beta * S * I / N),
    recovery = flow(from = "I", to = "R", rate = gamma * I)
  ),
  sums = list(N = c("S", "I", "R")),
  params = c("beta", "gamma")
)
```

### Stratify by 3 age groups

```{r stratify-3}
sir3 <- sf_stratify(sir, c("young", "adult", "old"),
  flow_types = c(infection = "disease", recovery = "disease"))
```

### Scalar vs array code

With the standard `sf_to_odin()`, each stratum gets its own named variables:

```{r scalar-code}
scalar_code <- sf_to_odin(sir3, type = "ode")
cat(scalar_code)
```

The code grows linearly with the number of strata. With `sf_to_odin_array()`,
we get compact array-indexed code:

```{r array-code}
array_code <- sf_to_odin_array(sir3, type = "ode")
cat(array_code)
```

The array code uses `dim()` declarations and `[]` indexing — odin2 automatically
loops over strata internally. This is much more concise and readable.

### Code size comparison

```{r size-compare}
cat(sprintf("Scalar code: %d characters\n", nchar(scalar_code)))
cat(sprintf("Array code:  %d characters\n", nchar(array_code)))
cat(sprintf("Ratio:       %.1fx smaller\n", nchar(scalar_code) / nchar(array_code)))
```

### Compile and simulate

```{r simulate-3}
gen <- sf_to_odin_array_system(sir3, type = "ode")
sys <- dust_system_create(gen(), list(
  beta = 0.3, gamma = 0.1,
  S0 = c(300, 500, 200),
  I0 = c(5, 3, 2),
  R0 = c(0, 0, 0)
))
dust_system_set_state_initial(sys)
t <- seq(0, 200)
res <- dust_system_simulate(sys, t)
st <- dust_unpack_state(sys, res)
```

```{r plot-3, fig.cap = "Age-stratified SIR (3 groups, array codegen)"}
par(mfrow = c(1, 3))
cols <- c("steelblue", "tomato", "seagreen")
strata_names <- c("Young", "Adult", "Old")

for (comp in c("S", "I", "R")) {
  matplot(t, t(st[[comp]]), type = "l", lty = 1, col = cols, lwd = 2,
    xlab = "Time", ylab = comp,
    main = paste(comp, "by age group"))
  legend("right", strata_names, col = cols, lty = 1, lwd = 2, cex = 0.8)
}
```

## Scaling to many strata

The real advantage of array codegen shows when scaling to many strata. Let's try
10 groups:

```{r scale-10}
sir10 <- sf_stratify(sir, paste0("g", 1:10),
  flow_types = c(infection = "disease", recovery = "disease"))

scalar10 <- sf_to_odin(sir10, type = "ode")
array10 <- sf_to_odin_array(sir10, type = "ode")

cat(sprintf("10 strata — scalar: %d chars, array: %d chars (%.1fx smaller)\n",
  nchar(scalar10), nchar(array10), nchar(scalar10) / nchar(array10)))
```

```{r simulate-10}
gen10 <- sf_to_odin_array_system(sir10, type = "ode")
sys10 <- dust_system_create(gen10(), list(
  beta = 0.3, gamma = 0.1,
  S0 = c(200, rep(100, 9)),
  I0 = c(10, rep(0, 9)),
  R0 = rep(0, 10)
))
dust_system_set_state_initial(sys10)
res10 <- dust_system_simulate(sys10, seq(0, 150))
st10 <- dust_unpack_state(sys10, res10)
```

```{r plot-10, fig.cap = "10-group SIR — infections by group"}
cols10 <- rainbow(10)
matplot(seq(0, 150), t(st10$I), type = "l", lty = 1, col = cols10, lwd = 1.5,
  xlab = "Time", ylab = "Infected",
  main = "Infections across 10 strata")
legend("topright", paste0("g", 1:10), col = cols10, lty = 1, ncol = 2, cex = 0.7)
```

## Cross-validation: scalar vs array

An important property is that array and scalar codegen produce identical
numerical results. Let's verify:

```{r cross-validate}
sir2 <- sf_stratify(sir, c("a", "b"),
  flow_types = c(infection = "disease", recovery = "disease"))

# Scalar
gen_s <- sf_to_odin_system(sir2, type = "ode")
sys_s <- dust_system_create(gen_s(), list(
  beta = 0.3, gamma = 0.1,
  S_a0 = 800, S_b0 = 600,
  I_a0 = 10, I_b0 = 5,
  R_a0 = 0, R_b0 = 0
))
dust_system_set_state_initial(sys_s)
res_s <- dust_system_simulate(sys_s, seq(0, 100))
st_s <- dust_unpack_state(sys_s, res_s)

# Array
gen_a <- sf_to_odin_array_system(sir2, type = "ode")
sys_a <- dust_system_create(gen_a(), list(
  beta = 0.3, gamma = 0.1,
  S0 = c(800, 600),
  I0 = c(10, 5),
  R0 = c(0, 0)
))
dust_system_set_state_initial(sys_a)
res_a <- dust_system_simulate(sys_a, seq(0, 100))
st_a <- dust_unpack_state(sys_a, res_a)

# Compare
max_diff_S <- max(abs(c(st_s$S_a, st_s$S_b) - as.vector(st_a$S)))
max_diff_I <- max(abs(c(st_s$I_a, st_s$I_b) - as.vector(st_a$I)))
max_diff_R <- max(abs(c(st_s$R_a, st_s$R_b) - as.vector(st_a$R)))
cat(sprintf("Max absolute difference — S: %g, I: %g, R: %g\n",
  max_diff_S, max_diff_I, max_diff_R))
```

The results are numerically identical (within floating-point precision).

## Stochastic models

Array codegen also supports stochastic (discrete-time) models. The generated
code uses `Binomial()` draws with array indexing:

```{r stochastic-code}
stoch_code <- sf_to_odin_array(sir3, type = "stochastic")
cat(stoch_code)
```

```{r simulate-stochastic, fig.cap = "Stochastic age-stratified SIR"}
gen_stoch <- sf_to_odin_array_system(sir3, type = "stochastic")
sys_stoch <- dust_system_create(gen_stoch(), list(
  beta = 0.3, gamma = 0.1,
  S0 = c(300, 500, 200),
  I0 = c(5, 3, 2),
  R0 = c(0, 0, 0)
), n_particles = 1, dt = 0.25)
dust_system_set_state_initial(sys_stoch)

t_stoch <- seq(0, 200)
res_stoch <- dust_system_simulate(sys_stoch, t_stoch)
st_stoch <- dust_unpack_state(sys_stoch, res_stoch)

par(mfrow = c(1, 3))
for (comp in c("S", "I", "R")) {
  vals <- st_stoch[[comp]]
  if (length(dim(vals)) == 3) vals <- vals[,,1]
  matplot(t_stoch, t(vals), type = "l", lty = 1, col = cols, lwd = 2,
    xlab = "Time", ylab = comp,
    main = paste("Stochastic", comp))
  legend("right", strata_names, col = cols, lty = 1, lwd = 2, cex = 0.8)
}
```

## Cross-strata flows (aging)

When stratification includes aging or migration between strata, `sf_to_odin_array()`
handles cross-strata flows using `if/else` conditionals:

```{r aging-code}
sir_aging <- sf_stratify(sir, c("young", "old"),
  flow_types = c(infection = "disease", recovery = "disease"),
  cross_strata_flows = list(
    list(name = "aging", from = "young", to = "old",
         rate = quote(aging_rate), params = "aging_rate")
  ))

aging_code <- sf_to_odin_array(sir_aging, type = "ode")
cat(aging_code)
```

```{r simulate-aging, fig.cap = "SIR with aging between young and old groups"}
gen_aging <- sf_to_odin_array_system(sir_aging, type = "ode")
sys_aging <- dust_system_create(gen_aging(), list(
  beta = 0.3, gamma = 0.1, aging_rate = 0.02,
  S0 = c(800, 200),
  I0 = c(10, 5),
  R0 = c(0, 0)
))
dust_system_set_state_initial(sys_aging)

t_age <- seq(0, 200)
res_age <- dust_system_simulate(sys_aging, t_age)
st_age <- dust_unpack_state(sys_aging, res_age)

par(mfrow = c(1, 2))
plot(t_age, st_age$I[1,], type = "l", col = "steelblue", lwd = 2,
  ylim = range(st_age$I), xlab = "Time", ylab = "Infected",
  main = "Infections by age group")
lines(t_age, st_age$I[2,], col = "tomato", lwd = 2)
legend("topright", c("Young", "Old"), col = c("steelblue", "tomato"),
  lty = 1, lwd = 2)

total <- colSums(st_age$S) + colSums(st_age$I) + colSums(st_age$R)
plot(t_age, total, type = "l", col = "black", lwd = 2,
  xlab = "Time", ylab = "Total population",
  main = "Population conservation")
```

## Summary

| Feature | `sf_to_odin()` (scalar) | `sf_to_odin_array()` |
|---------|------------------------|---------------------|
| Code size | O(n_strata) | O(1) |
| Readability | Decreases with strata | Constant |
| Cross-strata flows | Named variables | `if/else` conditionals |
| Contact matrices | Named parameters | 2D array parameters |
| ODE support | ✓ | ✓ |
| Stochastic support | ✓ | ✓ |
| Max practical strata | ~5–10 | 50+ |

Use `sf_to_odin()` for small models (2–3 strata) where named variables are clearer.
Use `sf_to_odin_array()` for larger models where compact code and scalability matter.
