---
title: "Compositional modeling"
author: "Simon Frost"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Compositional modeling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

The central idea of algebraicodin is **compositional modeling**: build
complex models by assembling reusable components. Instead of writing a
monolithic SIR model, we define *infection* and *recovery* as separate
**open Petri nets**, then compose them by declaring how they share
species.

This approach mirrors the mathematical framework of
[AlgebraicPetri.jl](https://algebraicjulia.github.io/AlgebraicPetri.jl/)
and uses **undirected wiring diagrams** (UWDs) from applied category
theory to specify the composition pattern.

```{r setup}
library(algebraicodin)
```

## Building blocks

An **open Petri net** is a Petri net with designated *legs*—species that
are exposed for composition. The package provides helpers for common
epidemiological transitions:

```{r blocks}
# Infection: S + I → I + I (frequency-dependent transmission)
infection <- exposure_petri("S", "I", "I", "inf")

# Recovery: I → R (spontaneous transition)
recovery <- spontaneous_petri("I", "R", "rec")
```

Each component is self-contained with its own species and transitions.
The legs expose species that will be identified during composition.

```{r blocks-diagram, fig.show = "hold"}
plot_petri(infection)
plot_petri(recovery)
```

## Composing with the `%o%` operator

The simplest way to compose two open Petri nets is the `%o%` operator,
which automatically identifies species with matching names:

```{r compose-op}
sir_composed <- infection %o% recovery
species_names(apex(sir_composed))
transition_names(apex(sir_composed))
```

Species "I" appears in both components, so it gets identified (shared).
The result is the standard SIR model.

```{r compose-diagram}
plot_petri(sir_composed)
```

```{r compose-odin}
sir_code <- pn_to_odin(apex(sir_composed), "ode")
cat(sir_code)
```

## Composing with wiring diagrams

For more control, define an explicit **undirected wiring diagram** (UWD)
and use `oapply()`. The UWD specifies *junctions* (shared species) and
*boxes* (components):

```{r uwd-sir}
w_sir <- catlab::uwd(
  outer = c("S", "I", "R"),
  infection = c("S", "I"),
  recovery  = c("I", "R")
)
sir_via_uwd <- oapply(w_sir, list(infection, recovery))
species_names(apex(sir_via_uwd))
```

The UWD approach is equivalent to `%o%` for two components but
generalises naturally to more complex diagrams.

```{r uwd-sir-diagram}
plot_uwd(w_sir)
```

## Extending to SEIR

Adding an exposed compartment requires a third component—*progression*
from E to I:

```{r seir}
# Exposure: S + I → E + I (infectious contact produces exposed, not infectious)
exposure <- exposure_petri("S", "I", "E", "exp")

# Progression: E → I
progression <- spontaneous_petri("E", "I", "prog")

# Compose with explicit UWD
w_seir <- catlab::uwd(
  outer    = c("S", "E", "I", "R"),
  exposure = c("S", "I", "E"),
  progression  = c("E", "I"),
  recovery = c("I", "R")
)
seir <- oapply(w_seir, list(exposure, progression, recovery))
species_names(apex(seir))
transition_names(apex(seir))
```

The wiring diagram shows how the three components share species:

```{r seir-diagram, fig.show = "hold"}
plot_uwd(w_seir)
plot_petri(seir)
```

```{r seir-odin}
cat(pn_to_odin(apex(seir), "ode"))
```

## Comparison: composed vs hand-written SEIR

The hand-written odin2 SEIR model:

```{r seir-handwritten}
seir_handwritten <- "
exp <- parameter()
prog <- parameter()
rec <- parameter()
S0 <- parameter()
E0 <- parameter()
I0 <- parameter()
R0 <- parameter()
initial(S) <- S0
initial(E) <- E0
initial(I) <- I0
initial(R) <- R0
rate_exp <- exp * S * I
rate_prog <- prog * E
rate_rec <- rec * I
deriv(S) <- -rate_exp
deriv(E) <- rate_exp - rate_prog
deriv(I) <- rate_prog - rate_rec
deriv(R) <- rate_rec
"
```

Let us run both and verify they produce identical trajectories:

```{r seir-compare}
# Algebraic version
gen_alg <- odin2::odin(pn_to_odin(apex(seir), "ode"))
pars <- list(exp = 0.001, prog = 0.2, rec = 0.1,
             S0 = 990, E0 = 0, I0 = 10, R0 = 0)
sys_alg <- dust2::dust_system_create(gen_alg, pars, n_particles = 1)
dust2::dust_system_set_state_initial(sys_alg)
t <- seq(0, 100, by = 0.5)
y_alg <- dust2::dust_system_simulate(sys_alg, t)

# Hand-written version
gen_hw <- odin2::odin(seir_handwritten)
sys_hw <- dust2::dust_system_create(gen_hw, pars, n_particles = 1)
dust2::dust_system_set_state_initial(sys_hw)
y_hw <- dust2::dust_system_simulate(sys_hw, t)

cat("Max absolute difference:", max(abs(y_alg - y_hw)), "\n")
```

The outputs match exactly—both compile to the same C++ system.

## Adding vaccination: SIRV

We can extend SIR with a vaccination transition (S → R at rate `vac`):

```{r sirv}
vaccination <- spontaneous_petri("S", "R", "vac")

w_sirv <- catlab::uwd(
  outer      = c("S", "I", "R"),
  infection  = c("S", "I"),
  recovery   = c("I", "R"),
  vaccination = c("S", "R")
)
sirv <- oapply(w_sirv, list(infection, recovery, vaccination))
cat(pn_to_odin(apex(sirv), "ode"))
```

```{r sirv-diagram, fig.show = "hold"}
plot_uwd(w_sirv)
plot_petri(sirv)
```

```{r sirv-run, fig.width = 6, fig.height = 4}
gen <- odin2::odin(pn_to_odin(apex(sirv), "ode"))
pars <- list(inf = 0.0005, rec = 0.25, vac = 0.05,
             S0 = 990, I0 = 10, R0 = 0)
sys <- dust2::dust_system_create(gen, pars, n_particles = 1)
dust2::dust_system_set_state_initial(sys)
y <- dust2::dust_system_simulate(sys, t)

df <- data.frame(
  time = rep(t, 3),
  value = c(y[1, ], y[2, ], y[3, ]),
  compartment = rep(c("S", "I", "R"), each = length(t))
)
library(ggplot2)
ggplot(df, aes(time, value, colour = compartment)) +
  geom_line(linewidth = 0.8) +
  labs(title = "SIRV model (composed from 3 components)",
       x = "Time", y = "Population", colour = "Compartment") +
  theme_minimal()
```

## Using the epidemiology dictionary

For convenience, `epi_dict()` provides pre-built components for common
transitions:

```{r epi-dict}
dict <- epi_dict()
names(dict)
```

You can compose a model directly from the dictionary using `compose_epi()`:

```{r compose-epi}
w <- catlab::uwd(
  outer     = c("S", "I", "R"),
  infection = c("S", "I"),
  recovery  = c("I", "R")
)
sir <- compose_epi(w, dict, c("infection", "recovery"))
species_names(apex(sir))
```

## Composing more than two components

The `compose()` function chains multiple `%o%` operations:

```{r compose-multi}
sirh <- compose(
  exposure_petri("S", "I", "I", "inf"),
  spontaneous_petri("I", "R", "rec"),
  spontaneous_petri("I", "H", "hosp"),
  spontaneous_petri("H", "R", "hosp_rec")
)
species_names(apex(sirh))
transition_names(apex(sirh))
```

## Summary

| Method | Syntax | Best for |
|--------|--------|----------|
| `%o%` | `a %o% b` | Quick binary composition |
| `oapply()` | `oapply(uwd, components)` | Full control via wiring diagrams |
| `compose()` | `compose(a, b, c)` | Chaining multiple components |
| `compose_epi()` | `compose_epi(uwd, dict, names)` | Dictionary-based workflows |

The compositional approach ensures that every shared species is correctly
identified and that stoichiometry is preserved across components. For
stratification by age, risk, or vaccination status, see
`vignette("stratification")`.
