---
title: "Model examples"
author: "Simon Frost"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Model examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
set.seed(42)
```

## Introduction

This vignette demonstrates a variety of epidemiological models built with
`algebraicodin`, showcasing Petri net, stock-flow, and compositional
(AlgebraicDynamics-style) approaches. Each model is generated as odin2
code, compiled, and simulated.

```{r setup, message=FALSE, warning=FALSE}
library(algebraicodin)
library(catlab)
library(ggplot2)
library(odin2)
library(dust2)
```

## 1. SIRS model (waning immunity)

The SIRS model extends SIR with waning immunity: recovered individuals
return to the susceptible class.

### Petri net approach

```{r sirs-petri}
sirs_pn <- labelled_petri_net(
  c("S", "I", "R"),
  infection = c("S", "I") %=>% c("I", "I"),
  recovery = "I" %=>% "R",
  waning = "R" %=>% "S"
)

code <- pn_to_odin(sirs_pn)
cat(code)
```

```{r sirs-simulate, message=FALSE}
gen <- pn_to_odin_system(sirs_pn)
sys <- dust_system_create(gen(), list(
  S0 = 990, I0 = 10, R0 = 0,
  infection = 0.3, recovery = 0.1, waning = 0.01
))
dust_system_set_state_initial(sys)
times <- seq(0, 500, by = 1)
res <- dust_system_simulate(sys, times)
state <- dust_unpack_state(sys, res)

df <- data.frame(
  time = rep(times, 3),
  value = c(state$S, state$I, state$R),
  compartment = rep(c("S", "I", "R"), each = length(times))
)

ggplot(df, aes(x = time, y = value, colour = compartment)) +
  geom_line(linewidth = 0.8) +
  labs(title = "SIRS model (Petri net)", x = "Time", y = "Count") +
  theme_minimal()
```

### Stock-flow approach

```{r sirs-sf}
sirs_sf <- stock_and_flow(
  stocks = c("S", "I", "R"),
  flows = list(
    infection = flow(from = "S", to = "I", rate = beta * S * I / N),
    recovery = flow(from = "I", to = "R", rate = gamma * I),
    waning = flow(from = "R", to = "S", rate = omega * R)
  ),
  sums = list(N = c("S", "I", "R")),
  params = c("beta", "gamma", "omega")
)

plot_stock_flow(sirs_sf)
```

```{r sirs-sf-simulate, message=FALSE}
gen <- sf_to_odin_system(sirs_sf, type = "ode")
sys <- dust_system_create(gen(), list(
  beta = 0.3, gamma = 0.1, omega = 0.01,
  S0 = 990, I0 = 10, R0 = 0
))
dust_system_set_state_initial(sys)
res <- dust_system_simulate(sys, times)
state <- dust_unpack_state(sys, res)

df <- data.frame(
  time = rep(times, 3),
  value = c(state$S, state$I, state$R),
  compartment = rep(c("S", "I", "R"), each = length(times))
)

ggplot(df, aes(x = time, y = value, colour = compartment)) +
  geom_line(linewidth = 0.8) +
  labs(title = "SIRS model (stock-flow)", x = "Time", y = "Count") +
  theme_minimal()
```

## 2. SIR with demographics (births and deaths)

An open-population SIR model where the total population is maintained
by births and natural deaths.

```{r sir-demog}
sir_demog <- stock_and_flow(
  stocks = c("S", "I", "R"),
  flows = list(
    birth = flow(from = NULL, to = "S", rate = mu * N),
    infection = flow(from = "S", to = "I", rate = beta * S * I / N),
    recovery = flow(from = "I", to = "R", rate = gamma * I),
    death_S = flow(from = "S", to = NULL, rate = mu * S),
    death_I = flow(from = "I", to = NULL, rate = mu * I),
    death_R = flow(from = "R", to = NULL, rate = mu * R)
  ),
  sums = list(N = c("S", "I", "R")),
  params = c("beta", "gamma", "mu")
)

plot_stock_flow(sir_demog)
```

```{r sir-demog-simulate, message=FALSE}
gen <- sf_to_odin_system(sir_demog, type = "ode")
sys <- dust_system_create(gen(), list(
  beta = 0.4, gamma = 0.1, mu = 0.01,
  S0 = 990, I0 = 10, R0 = 0
))
dust_system_set_state_initial(sys)
times <- seq(0, 500, by = 1)
res <- dust_system_simulate(sys, times)
state <- dust_unpack_state(sys, res)

df <- data.frame(
  time = rep(times, 3),
  value = c(state$S, state$I, state$R),
  compartment = rep(c("S", "I", "R"), each = length(times))
)

ggplot(df, aes(x = time, y = value, colour = compartment)) +
  geom_line(linewidth = 0.8) +
  labs(title = "SIR with demographics", x = "Time", y = "Count") +
  theme_minimal()
```

## 3. SEIR model

The SEIR model includes an exposed (latent) class between susceptible
and infectious.

```{r seir}
seir_pn <- labelled_petri_net(
  c("S", "E", "I", "R"),
  exposure = c("S", "I") %=>% c("E", "I"),
  infection = "E" %=>% "I",
  recovery = "I" %=>% "R"
)

code <- pn_to_odin(seir_pn)
cat(code)
```

```{r seir-simulate, message=FALSE}
gen <- pn_to_odin_system(seir_pn)
sys <- dust_system_create(gen(), list(
  S0 = 990, E0 = 0, I0 = 10, R0 = 0,
  exposure = 0.4, infection = 0.2, recovery = 0.1
))
dust_system_set_state_initial(sys)
times <- seq(0, 200, by = 0.5)
res <- dust_system_simulate(sys, times)
state <- dust_unpack_state(sys, res)

df <- data.frame(
  time = rep(times, 4),
  value = c(state$S, state$E, state$I, state$R),
  compartment = rep(c("S", "E", "I", "R"), each = length(times))
)

ggplot(df, aes(x = time, y = value, colour = compartment)) +
  geom_line(linewidth = 0.8) +
  labs(title = "SEIR model", x = "Time", y = "Count") +
  theme_minimal()
```

## 4. Compositional SIR using resource sharers

The same SIR model can be built by composing independent sub-models:

```{r composed-sir}
infection_rs <- continuous_resource_sharer(
  state_names = c("S", "I"),
  dynamics_expr = list(S = quote(-beta * S * I), I = quote(beta * S * I)),
  params = c("beta")
)

recovery_rs <- continuous_resource_sharer(
  state_names = c("I", "R"),
  dynamics_expr = list(I = quote(-gamma * I), R = quote(gamma * I)),
  params = c("gamma")
)

sir_uwd <- uwd(
  outer = c("s", "i", "r"),
  infection = c("s", "i"),
  recovery = c("i", "r")
)

sir_composed <- oapply_dynam(sir_uwd, list(
  infection = infection_rs,
  recovery = recovery_rs
))

code <- rs_to_odin(sir_composed, type = "ode",
  initial = c(S = 990, I = 10, R = 0))
cat(code)
```

```{r composed-simulate, message=FALSE}
gen_fn <- function() odin2::odin(code)
sys <- dust_system_create(gen_fn(), list(
  beta = 0.3, gamma = 0.1,
  S0 = 990, I0 = 10, R0 = 0
))
dust_system_set_state_initial(sys)
times <- seq(0, 150, by = 0.5)
res <- dust_system_simulate(sys, times)
state <- dust_unpack_state(sys, res)

df <- data.frame(
  time = rep(times, 3),
  value = c(state$S, state$I, state$R),
  compartment = rep(c("S", "I", "R"), each = length(times))
)

ggplot(df, aes(x = time, y = value, colour = compartment)) +
  geom_line(linewidth = 0.8) +
  labs(title = "SIR via composition (resource sharers)", x = "Time", y = "Count") +
  theme_minimal()
```

## 5. Stochastic SIR (Petri net)

Using odin2's stochastic generation for a discrete-time stochastic SIR:

```{r stochastic-sir}
sir_pn <- labelled_petri_net(
  c("S", "I", "R"),
  infection = c("S", "I") %=>% c("I", "I"),
  recovery = "I" %=>% "R"
)

stoch_code <- pn_to_odin(sir_pn, type = "stochastic")
cat(stoch_code)
```

```{r stochastic-simulate, message=FALSE}
gen <- pn_to_odin_system(sir_pn, type = "stochastic")
sys <- dust_system_create(gen(), list(
  S0 = 990, I0 = 10, R0 = 0,
  infection = 0.3, recovery = 0.1, dt = 0.1
), n_particles = 10)
dust_system_set_state_initial(sys)
times <- seq(0, 150, by = 1)
res <- dust_system_simulate(sys, times)
state <- dust_unpack_state(sys, res)

# Plot multiple particles for I
df_list <- lapply(1:10, function(p) {
  data.frame(time = times, I = state$I[p, ], particle = factor(p))
})
df <- do.call(rbind, df_list)

ggplot(df, aes(x = time, y = I, group = particle, colour = particle)) +
  geom_line(alpha = 0.6) +
  labs(title = "Stochastic SIR (10 particles)", x = "Time", y = "Infected") +
  theme_minimal() +
  theme(legend.position = "none")
```

## 6. Age-stratified SIR (stock-flow stratification)

Using the stratification machinery:

```{r age-sir}
sir_sf <- stock_and_flow(
  stocks = c("S", "I", "R"),
  flows = list(
    infection = flow(from = "S", to = "I", rate = beta * S * I / N),
    recovery = flow(from = "I", to = "R", rate = gamma * I)
  ),
  sums = list(N = c("S", "I", "R")),
  params = c("beta", "gamma")
)

sir_age <- sf_stratify(sir_sf, c("young", "old"),
  flow_types = c(infection = "disease", recovery = "disease"),
  cross_strata_flows = list(
    list(name = "aging", from = "young", to = "old",
         rate = quote(aging_rate * young), params = c("aging_rate"))
  )
)

plot_stock_flow(sir_age)
```

```{r age-sir-simulate, message=FALSE}
gen <- sf_to_odin_system(sir_age, type = "ode")
sys <- dust_system_create(gen(), list(
  beta = 0.3, gamma = 0.1, aging_rate = 0.02,
  S_young0 = 500, S_old0 = 400,
  I_young0 = 5, I_old0 = 5,
  R_young0 = 0, R_old0 = 0
))
dust_system_set_state_initial(sys)
times <- seq(0, 200, by = 0.5)
res <- dust_system_simulate(sys, times)
state <- dust_unpack_state(sys, res)

df <- data.frame(
  time = rep(times, 4),
  value = c(state$I_young, state$I_old, state$S_young, state$S_old),
  compartment = rep(c("I_young", "I_old", "S_young", "S_old"),
                    each = length(times))
)

ggplot(df, aes(x = time, y = value, colour = compartment)) +
  geom_line(linewidth = 0.8) +
  labs(title = "Age-stratified SIR", x = "Time", y = "Count") +
  theme_minimal()
```

## 7. Lotka-Volterra predator-prey

Stock-flow models aren't limited to epidemiology. Here's the classic
Lotka-Volterra predator-prey system:

```{r lotka-volterra}
lv <- stock_and_flow(
  stocks = c("prey", "predator"),
  flows = list(
    prey_growth = flow(from = NULL, to = "prey", rate = alpha * prey),
    predation = flow(from = "prey", to = NULL, rate = beta * prey * predator),
    predator_growth = flow(from = NULL, to = "predator",
                           rate = delta * prey * predator),
    predator_death = flow(from = "predator", to = NULL, rate = gamma * predator)
  ),
  params = c("alpha", "beta", "delta", "gamma")
)

plot_stock_flow(lv)
```

```{r lv-simulate, message=FALSE}
gen <- sf_to_odin_system(lv, type = "ode")
sys <- dust_system_create(gen(), list(
  alpha = 1.1, beta = 0.4, delta = 0.1, gamma = 0.4,
  prey0 = 10, predator0 = 10
))
dust_system_set_state_initial(sys)
times <- seq(0, 50, by = 0.05)
res <- dust_system_simulate(sys, times)
state <- dust_unpack_state(sys, res)

# Phase plot
df_phase <- data.frame(prey = state$prey, predator = state$predator)
ggplot(df_phase, aes(x = prey, y = predator)) +
  geom_path(colour = "steelblue", linewidth = 0.5) +
  geom_point(data = df_phase[1, ], colour = "red", size = 3) +
  labs(title = "Lotka-Volterra phase portrait", x = "Prey", y = "Predator") +
  theme_minimal()
```

## Summary

| Model | Approach | Key feature |
|-------|----------|-------------|
| SIRS | Petri net + stock-flow | Waning immunity |
| SIR demographics | Stock-flow | Open population |
| SEIR | Petri net | Latent class |
| Composed SIR | Resource sharers | Compositional |
| Stochastic SIR | Petri net | Multiple particles |
| Age-stratified SIR | Stock-flow stratification | Cross-strata aging |
| Lotka-Volterra | Stock-flow | Non-epi application |

All models are generated as odin2 code, compiled to C++ by dust2, and
simulated efficiently.
