---
title: "Fitting models to data"
author: "Simon Frost"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Fitting models to data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
set.seed(42)
```

## Introduction

A key advantage of composing models algebraically is that the same model
definition used for simulation can also be used for **data fitting**.
`algebraicodin` integrates with dust2's particle filter and monty's MCMC
infrastructure to enable Bayesian inference on compartmental models.

This vignette walks through:

1. Generating synthetic epidemic data
2. Adding an observation model to a Petri net
3. Creating a particle filter
4. Running MCMC to recover parameters
5. Plotting diagnostics and posterior distributions

```{r setup, message=FALSE, warning=FALSE}
library(algebraicodin)
library(catlab)
library(ggplot2)
library(odin2)
library(dust2)
library(monty)
```

## Defining the model

We start with a simple SIR Petri net:

```{r model}
sir <- labelled_petri_net(
  c("S", "I", "R"),
  inf = c("S", "I") %=>% c("I", "I"),
  rec = "I" %=>% "R"
)
```

## Generating synthetic data

First, simulate an epidemic using the stochastic model to create
"observed" incidence data:

```{r synthetic-data}
gen <- pn_to_odin_system(sir, type = "stochastic")

# True parameters
true_pars <- list(inf = 3e-6, rec = 0.1,
                  S0 = 9990, I0 = 10, R0 = 0)

sys <- dust_system_create(gen(), true_pars,
                          n_particles = 1, dt = 1, seed = 1)
dust_system_set_state_initial(sys)
times <- seq(0, 100)
res <- dust_system_simulate(sys, times)
state <- dust_unpack_state(sys, res)

# Extract "true" incidence (new infections per day)
new_inf <- c(0, pmax(0, -diff(state$S)))

# Observe with Poisson noise at weekly intervals
obs_times <- seq(7, 98, by = 7)
obs_cases <- sapply(obs_times, function(t) {
  idx <- which(times >= (t - 6) & times <= t)
  rpois(1, lambda = max(1, sum(new_inf[idx])))
})

data <- data.frame(time = obs_times, cases = obs_cases)

ggplot(data, aes(x = time, y = cases)) +
  geom_col(fill = "steelblue", alpha = 0.7) +
  labs(title = "Synthetic weekly case data",
       x = "Day", y = "Weekly cases") +
  theme_minimal()
```

## Adding an observation model

The `observe()` function specifies how model quantities relate to data.
For incidence data, we track the `inf` transition and compare with a
Poisson distribution. Adding a noise parameter prevents zero-lambda
issues:

```{r compare}
compare_spec <- list(
  cases = observe("Poisson", transition = "inf", noise = "exp_noise")
)

# See generated code
code <- pn_to_odin(sir, type = "stochastic", compare = compare_spec)
cat(code)
```

The key additions are:
- **Incidence tracking**: `incidence_inf` accumulates new infections,
  resetting each timestep via `zero_every`
- **Noise**: `1 / exp_noise` adds a tiny positive floor
- **Comparison**: `cases ~ Poisson(noise_cases)` defines the likelihood

## Creating a particle filter

```{r filter}
filter <- create_filter(sir, data,
  compare = compare_spec,
  type = "stochastic",
  n_particles = 100)
```

Test the likelihood at the true parameters:

```{r likelihood}
pars_test <- c(true_pars, list(exp_noise = 1e6))
ll <- dust_likelihood_run(filter, pars_test)
cat("Log-likelihood at true params:", ll, "\n")
```

## Setting up MCMC

### Prior distributions

```{r prior}
prior <- build_prior(
  inf = prior_uniform(min = 1e-7, max = 1e-4),
  rec = prior_uniform(min = 0.01, max = 0.5)
)
```

### Running the sampler

```{r mcmc, message=FALSE}
samples <- fit_mcmc(
  filter = filter,
  pars = c("inf", "rec"),
  fixed = list(S0 = 9990, I0 = 10, R0 = 0, exp_noise = 1e6),
  prior = prior,
  n_steps = 500,
  n_chains = 2,
  initial = c(3e-6, 0.1),
  vcv = diag(c(1e-14, 1e-4))
)
```

## Diagnostics

### Trace plots

```{r traces}
plot_traces(samples, pars = c("inf", "rec"), burnin = 100)
```

### Posterior distributions

```{r posteriors}
plot_posterior(samples, burnin = 100,
  true_values = list(inf = 3e-6, rec = 0.1))
```

### Parameter summary

```{r summary}
burnin <- 100
n_steps <- dim(samples$pars)[2]

for (i in 1:2) {
  nm <- c("inf", "rec")[i]
  vals <- as.vector(samples$pars[i, (burnin+1):n_steps, ])
  cat(sprintf("%s: mean=%.2e, median=%.2e, 95%% CI=[%.2e, %.2e]\n",
    nm, mean(vals), median(vals),
    quantile(vals, 0.025), quantile(vals, 0.975)))
}
```

## Stock-flow model fitting

The same workflow applies to stock-flow models:

```{r sf-fitting, message=FALSE}
sir_sf <- stock_and_flow(
  stocks = c("S", "I", "R"),
  flows = list(
    infection = flow(from = "S", to = "I", rate = beta * S * I / N),
    recovery = flow(from = "I", to = "R", rate = gamma * I)
  ),
  sums = list(N = c("S", "I", "R")),
  params = c("beta", "gamma")
)

# Create filter with flow-based incidence tracking
filter_sf <- create_filter(sir_sf, data,
  compare = list(
    cases = observe("Poisson", flow = "infection", noise = "exp_noise")
  ),
  type = "stochastic",
  n_particles = 100)

# Check likelihood
ll_sf <- dust_likelihood_run(filter_sf, list(
  beta = 0.03, gamma = 0.1,
  S0 = 9990, I0 = 10, R0 = 0, exp_noise = 1e6
))
cat("Stock-flow log-likelihood:", ll_sf, "\n")
```

## Prevalence data

For data that measures current state (not incidence), use
`state` instead of `transition`:

```{r prevalence}
code_prev <- pn_to_odin(sir, type = "stochastic",
  compare = list(
    prevalence = observe("Normal", state = "I", sd = "obs_sd")
  ))
cat(code_prev)
```

## Composed model fitting

Algebraically composed models work seamlessly:

```{r composed, message=FALSE}
# Compose SIR from building blocks
dict <- epi_dict()
sir_uwd <- uwd(
  outer = c("s", "i", "r"),
  infection = c("s", "i"),
  recovery = c("i", "r")
)
sir_composed <- apex(oapply(sir_uwd, list(dict$infection, dict$recovery)))

# Generate with comparison — works identically
code_comp <- pn_to_odin(sir_composed, type = "stochastic",
  compare = list(cases = observe("Poisson", transition = "inf")))
cat(code_comp)
```

## Summary

| Function | Purpose |
|----------|---------|
| `observe()` | Define observation model |
| `create_filter()` | Build dust2 particle filter |
| `fit_mcmc()` | Run MCMC inference |
| `build_prior()` | Create prior from specs |
| `plot_traces()` | MCMC trace plots |
| `plot_posterior()` | Posterior density plots |
| `plot_fit()` | Model vs data comparison |

Key observation types:
- **Incidence**: `observe("Poisson", transition = "inf")` — tracks new
  events per timestep
- **Prevalence**: `observe("Normal", state = "I", sd = "sigma")` —
  observes current state
- **Custom**: `observe("Poisson", expr = "mu * I")` — arbitrary
  expression
