---
title: "Model types: ODE, DDE, stochastic, and discrete"
author: "Simon Frost"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Model types: ODE, DDE, stochastic, and discrete}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
set.seed(42)
```

## Introduction

A single Petri net can generate odin2 code for four different model types,
following the framework of
[AlgebraicDynamics.jl](https://algebraicjulia.github.io/AlgebraicDynamics.jl/).
Each type uses the same stoichiometry but differs in how the dynamics
are computed.

| Type | odin2 keyword | Time | Randomness |
|------|---------------|------|------------|
| **ODE** | `deriv()` | Continuous | Deterministic |
| **DDE** | `deriv()` + `delay()` | Continuous | Deterministic |
| **Stochastic** | `update()` + `Binomial()` | Discrete | Stochastic |
| **Discrete** | `update()` | Discrete | Deterministic |

```{r setup}
library(algebraicodin)
```

We use the standard SIR model throughout:

```{r sir}
sir <- labelled_petri_net(
  c("S", "I", "R"),
  inf = c("S", "I") %=>% c("I", "I"),
  rec = "I" %=>% "R"
)
```

```{r sir-diagram}
plot_petri(sir)
```

## ODE (continuous deterministic)

The default model type. Uses `deriv()` to define continuous derivatives
solved by numerical integration:

```{r ode}
cat(pn_to_odin(sir, "ode"))
```

```{r ode-run, fig.width = 6, fig.height = 4}
gen <- odin2::odin(pn_to_odin(sir, "ode"))
pars <- list(inf = 0.0005, rec = 0.25, S0 = 990, I0 = 10, R0 = 0)
sys <- dust2::dust_system_create(gen, pars, n_particles = 1)
dust2::dust_system_set_state_initial(sys)
t_ode <- seq(0, 40, by = 0.1)
y_ode <- dust2::dust_system_simulate(sys, t_ode)
```

### Hand-written comparison

```{r ode-hw}
ode_handwritten <- "
inf <- parameter()
rec <- parameter()
S0 <- parameter()
I0 <- parameter()
R0 <- parameter()
initial(S) <- S0
initial(I) <- I0
initial(R) <- R0
deriv(S) <- -inf * S * I
deriv(I) <- inf * S * I - rec * I
deriv(R) <- rec * I
"
gen_hw <- odin2::odin(ode_handwritten)
sys_hw <- dust2::dust_system_create(gen_hw, pars, n_particles = 1)
dust2::dust_system_set_state_initial(sys_hw)
y_hw <- dust2::dust_system_simulate(sys_hw, t_ode)
cat("ODE max diff:", max(abs(y_ode - y_hw)), "\n")
```

## DDE (delay differential equations)

DDEs introduce delays into the dynamics. In epidemiology, this can model
a fixed incubation period or delayed immune response.

The `delays` argument specifies which transitions use delayed values and
the delay time (`tau`):

```{r dde}
dde_code <- pn_to_odin(sir, "dde", delays = list(
  inf = list(tau = 5, species = "I")
))
cat(dde_code)
```

The infection rate now uses `delay(I, 5)` instead of the current value of
I, meaning that infectiousness depends on the state 5 time units in the
past.

### Hand-written comparison

```{r dde-hw}
dde_handwritten <- "
inf <- parameter()
rec <- parameter()
S0 <- parameter()
I0 <- parameter()
R0 <- parameter()
initial(S) <- S0
initial(I) <- I0
initial(R) <- R0
I_delayed <- delay(I, 5)
deriv(S) <- -inf * S * I_delayed
deriv(I) <- inf * S * I_delayed - rec * I
deriv(R) <- rec * I
"
```

```{r dde-run, fig.width = 6, fig.height = 4}
gen_dde <- odin2::odin(dde_code)
pars_dde <- list(inf = 0.0005, rec = 0.25, S0 = 990, I0 = 10, R0 = 0)
sys_dde <- dust2::dust_system_create(gen_dde, pars_dde, n_particles = 1)
dust2::dust_system_set_state_initial(sys_dde)
y_dde <- dust2::dust_system_simulate(sys_dde, t_ode)

gen_hw2 <- odin2::odin(dde_handwritten)
sys_hw2 <- dust2::dust_system_create(gen_hw2, pars_dde, n_particles = 1)
dust2::dust_system_set_state_initial(sys_hw2)
y_hw2 <- dust2::dust_system_simulate(sys_hw2, t_ode)
cat("DDE max diff:", max(abs(y_dde - y_hw2)), "\n")
```

```{r dde-plot, fig.width = 6, fig.height = 4}
library(ggplot2)
df <- data.frame(
  time = rep(t_ode, 2),
  I = c(y_ode[2, ], y_dde[2, ]),
  model = rep(c("ODE", "DDE (tau=5)"), each = length(t_ode))
)
ggplot(df, aes(time, I, colour = model)) +
  geom_line(linewidth = 0.8) +
  labs(title = "Effect of delay on infection dynamics",
       x = "Time", y = "Infectious (I)") +
  theme_minimal()
```

The delay shifts the epidemic peak later and reduces its height, as the
infection rate responds to historical rather than current infectiousness.

## Stochastic (discrete-time)

The stochastic model uses `Binomial()` draws for each transition,
modelling demographic stochasticity:

```{r stochastic}
cat(pn_to_odin(sir, "stochastic"))
```

Key features:
- Transition probabilities: $p = 1 - e^{-\text{rate} \cdot dt}$
- Draws: `Binomial(substrate, probability)`
- Update equations use draws instead of rates

### Hand-written comparison

```{r stoch-hw}
stoch_handwritten <- "
inf <- parameter()
rec <- parameter()
S0 <- parameter()
I0 <- parameter()
R0 <- parameter()
initial(S) <- S0
initial(I) <- I0
initial(R) <- R0
p_inf <- 1 - exp(-(inf * I) * dt)
n_inf <- Binomial(S, p_inf)
p_rec <- 1 - exp(-(rec) * dt)
n_rec <- Binomial(I, p_rec)
update(S) <- S - n_inf
update(I) <- I + n_inf - n_rec
update(R) <- R + n_rec
"
```

```{r stoch-run, fig.width = 6, fig.height = 4}
gen_stoch <- odin2::odin(pn_to_odin(sir, "stochastic"))
pars_stoch <- list(inf = 0.0005, rec = 0.25,
                   S0 = 990, I0 = 10, R0 = 0)
sys_stoch <- dust2::dust_system_create(gen_stoch, pars_stoch, n_particles = 10,
                                        seed = 42, dt = 1)
dust2::dust_system_set_state_initial(sys_stoch)
t_disc <- seq(0L, 40L, by = 1L)
y_stoch <- dust2::dust_system_simulate(sys_stoch, t_disc)

# Plot multiple realisations
dfs <- do.call(rbind, lapply(1:10, function(p) {
  data.frame(time = t_disc, I = y_stoch[2, p, ], particle = p)
}))
ggplot(dfs, aes(time, I, group = particle)) +
  geom_line(alpha = 0.5, colour = "steelblue") +
  geom_line(data = data.frame(time = t_ode, I = y_ode[2, ]),
            aes(group = NULL), colour = "red", linewidth = 1) +
  labs(title = "Stochastic SIR (10 realisations) vs ODE (red)",
       x = "Time", y = "Infectious (I)") +
  theme_minimal()
```

## Discrete deterministic (Euler method)

The discrete deterministic model uses a forward Euler step, useful for
comparison with the stochastic model or when discrete-time dynamics are
desired:

```{r discrete}
cat(pn_to_odin(sir, "discrete"))
```

### Hand-written comparison

```{r disc-hw}
disc_handwritten <- "
inf <- parameter()
rec <- parameter()
S0 <- parameter()
I0 <- parameter()
R0 <- parameter()
initial(S) <- S0
initial(I) <- I0
initial(R) <- R0
rate_inf <- inf * S * I
rate_rec <- rec * I
update(S) <- S - dt * rate_inf
update(I) <- I + dt * rate_inf - dt * rate_rec
update(R) <- R + dt * rate_rec
"
gen_disc <- odin2::odin(pn_to_odin(sir, "discrete"))
gen_hw3 <- odin2::odin(disc_handwritten)

pars_disc <- list(inf = 0.0005, rec = 0.25,
                  S0 = 990, I0 = 10, R0 = 0)
sys_disc <- dust2::dust_system_create(gen_disc, pars_disc, n_particles = 1, dt = 1)
dust2::dust_system_set_state_initial(sys_disc)
t_int <- seq(0L, 40L, by = 1L)
y_disc <- dust2::dust_system_simulate(sys_disc, t_int)

sys_hw3 <- dust2::dust_system_create(gen_hw3, pars_disc, n_particles = 1, dt = 1)
dust2::dust_system_set_state_initial(sys_hw3)
y_hw3 <- dust2::dust_system_simulate(sys_hw3, t_int)
cat("Discrete max diff:", max(abs(y_disc - y_hw3)), "\n")
```

## ResourceSharer and Machine types

For programmatic manipulation of dynamical systems (rather than code
generation), algebraicodin provides `ResourceSharer` and `Machine`
classes following AlgebraicDynamics.jl:

```{r resource-sharer}
rs <- petri_to_continuous(sir)
rs@system_type
rs@nstates
rs@state_names
```

Evaluate the vector field at a given state:

```{r eval-dynamics}
u <- c(990, 10, 0)
p <- list(inf = 0.0005, rec = 0.25)
eval_dynamics(rs, u, p, t = 0)
```

## Euler approximation

Convert a continuous system to discrete via forward Euler:

```{r euler}
rs_disc <- euler_approx(rs, h = 0.1)
rs_disc@system_type

# One Euler step
u_next <- eval_dynamics(rs_disc, u, p, t = 0)
cat("u_next:", u_next, "\n")
cat("u + 0.1 * f(u):", u + 0.1 * eval_dynamics(rs, u, p, t = 0), "\n")
```

## All four types from one Petri net

```{r all-four, fig.width = 7, fig.height = 5}
# Run all four model types
run_model <- function(code, pars, times, type) {
  gen <- odin2::odin(code)
  is_discrete <- type %in% c("Stochastic", "Discrete")
  args <- list(generator = gen, pars = pars, n_particles = 1L,
               deterministic = (type != "Stochastic"), seed = 42L)
  if (is_discrete) args$dt <- 1
  sys <- do.call(dust2::dust_system_create, args)
  dust2::dust_system_set_state_initial(sys)
  y <- dust2::dust_system_simulate(sys, times)
  data.frame(time = times, I = y[2, ], type = type)
}

pars_cont <- list(inf = 0.0005, rec = 0.25, S0 = 990, I0 = 10, R0 = 0)
pars_disc <- pars_cont

dde_delays <- list(inf = list(tau = 5, species = "I"))

df_all <- rbind(
  run_model(pn_to_odin(sir, "ode"), pars_cont, t_ode, "ODE"),
  run_model(pn_to_odin(sir, "dde", delays = dde_delays), pars_cont, t_ode, "DDE"),
  run_model(pn_to_odin(sir, "stochastic"), pars_disc, t_disc, "Stochastic"),
  run_model(pn_to_odin(sir, "discrete"), pars_disc, t_disc, "Discrete")
)

ggplot(df_all, aes(time, I, colour = type)) +
  geom_line(linewidth = 0.8) +
  labs(title = "SIR model: four types from one Petri net",
       x = "Time", y = "Infectious (I)", colour = "Model type") +
  theme_minimal()
```

## Summary

| Type | `pn_to_odin()` | Use case |
|------|-------------|----------|
| `"ode"` | `deriv()` | Standard deterministic dynamics |
| `"dde"` | `deriv()` + `delay()` | Fixed delays (incubation, immunity) |
| `"stochastic"` | `update()` + `Binomial()` | Demographic stochasticity |
| `"discrete"` | `update()` (Euler) | Discrete-time deterministic |

The Petri net is the single source of truth. Changing the model type is a
one-line change to `pn_to_odin(sir, type = ...)`, and the stoichiometry
remains consistent across all four formulations.
