---
title: "Stratification with typed products"
author: "Simon Frost"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Stratification with typed products}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

Many epidemiological models stratify a base compartmental model by
attributes like age, risk group, or vaccination status. Doing this by hand
is error-prone: an SIR model stratified by 3 age groups has 9 species
and many transitions to track.

algebraicodin automates stratification using **typed Petri nets** and their
**typed product**, porting the approach from
[AlgebraicPetri.jl](https://algebraicjulia.github.io/AlgebraicPetri.jl/).
This is the categorical product in the slice category $\mathrm{Petri}/T$,
where $T$ is a *type system* (ontology) that specifies the allowed
transition types.

```{r setup}
library(algebraicodin)
```

## Type systems (ontologies)

A **type system** is a Petri net that defines abstract transition
patterns. The `infectious_ontology()` provides the standard type system
for infectious disease models:

```{r ontology}
ont <- infectious_ontology()
species_names(ont)
transition_names(ont)
```

The ontology has one species type (**Pop**) and three transition types:

| Type | Input → Output | Meaning |
|------|----------------|---------|
| **infect** | Pop, Pop → Pop, Pop | Frequency-dependent transmission |
| **disease** | Pop → Pop | Spontaneous transition (recovery, death, ...) |
| **strata** | Pop → Pop | Movement between strata (aging, migration, ...) |

## Typed Petri nets

A **typed Petri net** is a morphism $\phi: P \to T$ from a concrete Petri
net to the type system. Each species and transition in $P$ is assigned a
type in $T$.

```{r sir-typed}
sir <- labelled_petri_net(
  c("S", "I", "R"),
  inf = c("S", "I") %=>% c("I", "I"),
  rec = "I" %=>% "R"
)
sir_typed <- typed_petri(sir, ont,
  species_types    = c(S = "Pop", I = "Pop", R = "Pop"),
  transition_types = c(inf = "infect", rec = "disease")
)
sir_typed@species_map
sir_typed@transition_map
```

```{r sir-typed-diagram}
plot_petri(sir_typed)
```

All SIR species map to `Pop` (index 1). The `inf` transition maps to
`infect` (index 1) and `rec` maps to `disease` (index 2).

## Reflexive transitions

Before taking a typed product, each model must be augmented with
**reflexive transitions**—identity self-loops that allow the product to
pair "real" transitions in one model with "do-nothing" transitions in the
other.

For stratification by age, the SIR model needs reflexive `strata`
transitions (so that aging can be paired with each compartment):

```{r sir-refl}
sir_aug <- add_reflexives(sir_typed, list(
  S = "strata",
  I = "strata",
  R = "strata"
))
transition_names(sir_aug@pn)
```

The augmented model now has 5 transitions:

```{r sir-aug-diagram}
plot_petri(sir_aug)
```

## Defining strata

Age groups are modelled as a simple Petri net with `strata`-typed
transitions. For two age groups with aging:

```{r age-model}
age <- labelled_petri_net(
  c("Young", "Old"),
  aging = "Young" %=>% "Old"
)
age_typed <- typed_petri(age, ont,
  species_types    = c(Young = "Pop", Old = "Pop"),
  transition_types = c(aging = "strata")
)
```

The age model also needs reflexive transitions for `infect` and `disease`
so these processes can occur within each age group:

```{r age-refl}
age_aug <- add_reflexives(age_typed, list(
  Young = c("infect", "disease"),
  Old   = c("infect", "disease")
))
transition_names(age_aug@pn)
```

```{r age-aug-diagram}
plot_petri(age_aug)
```

## Taking the typed product

The **typed product** (or stratification) creates all valid pairs of
species and transitions. The `%x%` operator is shorthand for
`typed_product()`:

```{r stratify}
sir_age <- sir_aug %x% age_aug
species_names(sir_age)
transition_names(sir_age)
```

The stratified model has 6 species (S, I, R for each age group) and 7
transitions (within-group infection, cross-group infection, and recovery
for each group, plus age transitions).

```{r product-diagram}
plot_petri(sir_age)
```

The result has:

- **6 species**: S×{Young,Old}, I×{Young,Old}, R×{Young,Old}
- **7 transitions**: infection and recovery *within* each age group, plus
  aging *across* each compartment

## Generated odin2 code

```{r stratify-odin}
code <- pn_to_odin(sir_age, "ode")
cat(code)
```

Each transition rate uses mass-action kinetics within the appropriate
stratum. The aging transitions move individuals between age groups while
preserving their disease state.

## Comparison with hand-written code

Here is the equivalent hand-written odin2 model:

```{r handwritten-age}
handwritten <- "
inf_Young <- parameter()
inf_Old <- parameter()
rec_Young <- parameter()
rec_Old <- parameter()
aging_S <- parameter()
aging_I <- parameter()
aging_R <- parameter()

S_Young0 <- parameter()
S_Old0 <- parameter()
I_Young0 <- parameter()
I_Old0 <- parameter()
R_Young0 <- parameter()
R_Old0 <- parameter()

initial(S_Young) <- S_Young0
initial(S_Old) <- S_Old0
initial(I_Young) <- I_Young0
initial(I_Old) <- I_Old0
initial(R_Young) <- R_Young0
initial(R_Old) <- R_Old0

rate_inf_Young <- inf_Young * S_Young * I_Young
rate_inf_Old <- inf_Old * S_Old * I_Old
rate_rec_Young <- rec_Young * I_Young
rate_rec_Old <- rec_Old * I_Old
rate_aging_S <- aging_S * S_Young
rate_aging_I <- aging_I * I_Young
rate_aging_R <- aging_R * R_Young

deriv(S_Young) <- -rate_inf_Young - rate_aging_S
deriv(S_Old) <- -rate_inf_Old + rate_aging_S
deriv(I_Young) <- rate_inf_Young - rate_rec_Young - rate_aging_I
deriv(I_Old) <- rate_inf_Old - rate_rec_Old + rate_aging_I
deriv(R_Young) <- rate_rec_Young - rate_aging_R
deriv(R_Old) <- rate_rec_Old + rate_aging_R
"
```

Running both:

```{r compare-age, fig.width = 7, fig.height = 4}
pars <- list(
  inf_Young = 0.001, inf_Old = 0.0005,
  rec_Young = 0.25, rec_Old = 0.25,
  aging_S = 0.01, aging_I = 0.01, aging_R = 0.01,
  S_Young0 = 450, S_Old0 = 450,
  I_Young0 = 5, I_Old0 = 5,
  R_Young0 = 0, R_Old0 = 0
)
t <- seq(0, 80, by = 0.5)

gen_alg <- odin2::odin(code)
sys_alg <- dust2::dust_system_create(gen_alg, pars, n_particles = 1)
dust2::dust_system_set_state_initial(sys_alg)
y_alg <- dust2::dust_system_simulate(sys_alg, t)

gen_hw <- odin2::odin(handwritten)
sys_hw <- dust2::dust_system_create(gen_hw, pars, n_particles = 1)
dust2::dust_system_set_state_initial(sys_hw)
y_hw <- dust2::dust_system_simulate(sys_hw, t)

cat("Max absolute difference:", max(abs(y_alg - y_hw)), "\n")

# Plot
snames <- species_names(sir_age)
df <- data.frame(
  time = rep(t, length(snames)),
  value = as.vector(t(y_alg)),
  compartment = rep(snames, each = length(t))
)
library(ggplot2)
ggplot(df, aes(time, value, colour = compartment)) +
  geom_line(linewidth = 0.7) +
  labs(title = "Age-stratified SIR (algebraicodin → odin2)",
       x = "Time", y = "Population") +
  theme_minimal()
```

## Three age groups

Scaling to more strata is trivial. Here is SIR × 3 age groups with
pairwise aging:

```{r three-ages}
age3 <- labelled_petri_net(
  c("Young", "Middle", "Old"),
  aging_YM = "Young" %=>% "Middle",
  aging_MO = "Middle" %=>% "Old"
)
age3_typed <- typed_petri(age3, ont,
  species_types    = c(Young = "Pop", Middle = "Pop", Old = "Pop"),
  transition_types = c(aging_YM = "strata", aging_MO = "strata")
)
age3_aug <- add_reflexives(age3_typed, list(
  Young  = c("infect", "disease"),
  Middle = c("infect", "disease"),
  Old    = c("infect", "disease")
))

sir_age3 <- sir_aug %x% age3_aug
cat("Species:", species_names(sir_age3), "\n")
cat("Transitions:", transition_names(sir_age3), "\n")
cat("nparts: S =", acsets::nparts(sir_age3@pn, "S"),
    ", T =", acsets::nparts(sir_age3@pn, "T"), "\n")
```

9 species and 12 transitions—automatically generated with correct
stoichiometry.

## Risk group stratification

Risk groups follow the same pattern. Here we stratify SIR by two risk
levels without movement between groups (no strata transitions):

```{r risk}
risk <- labelled_petri_net(c("High", "Low"))
risk_typed <- typed_petri(risk, ont,
  species_types    = c(High = "Pop", Low = "Pop"),
  transition_types = setNames(character(0), character(0))
)
risk_aug <- add_reflexives(risk_typed, list(
  High = c("infect", "disease"),
  Low  = c("infect", "disease")
))

sir_risk <- sir_aug %x% risk_aug
species_names(sir_risk)
transition_names(sir_risk)
```

Without strata transitions, there is no movement between risk groups.
The infection and recovery happen independently within each group.

## The `stratify()` wrapper

`stratify()` is an alias for `typed_product()`:

```{r stratify-fn}
sir_age2 <- stratify(sir_aug, age_aug)
identical(species_names(sir_age2), species_names(sir_age))
```

## Stochastic stratified models

The stratified Petri net can generate any model type. Here is the
age-stratified SIR as a stochastic model:

```{r stochastic-strat}
cat(pn_to_odin(sir_age, "stochastic"))
```

## Summary

| Step | Function | Purpose |
|------|----------|---------|
| Define type system | `infectious_ontology()` | Declare allowed transition types |
| Assign types | `typed_petri()` | Create morphism PN → ontology |
| Add reflexives | `add_reflexives()` | Enable cross-model pairing |
| Stratify | `%x%` or `typed_product()` | Compute categorical product |
| Flatten | `flatten()` | Extract underlying PN from typed PN |
| Generate code | `pn_to_odin()` | Produce odin2 DSL for any model type |

The typed product guarantees structural correctness: every species pair
has consistent transitions, and the stoichiometry is preserved. Scaling
from 2 to 20 strata requires only changing the strata model—the base
disease model is unchanged.
